---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# sliopt: Single Link Indicator OPTimization for geospatial analysis

<!-- badges: start -->
[![R-CMD-check](https://github.com/chris31415926535/sliopt/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/chris31415926535/sliopt/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

Sliopt provides an algorithm to create optimized single-link indicators from one set of (non-self-overlapping) geometries to a different overlapping set of (non-self-overlapping) geometries. It requires you to have one "source of truth" for each set of geometries.

Sometimes you have data measured over one set of geometries, like census tracts, but you want to work with a different larger set of geometries, like school catchment areas. One approach is to map each smaller geometry to one and only one larger geometry using a "single link indicator," or "SLI." This will allow you to aggregate the values from the smaller regions up to the bigger regions.

If the two sets of geometries nest into each other perfectly, then it's trivial to create an SLI. But if the two geometry sets have borders that don't align, there is no single way to map each member of the first set to a member of the second set.

This package assumes that you are able to get at least one set of values that is reliable across all geometries. It then uses that set of "ground truths" to create an optimized SLI linking the two sets of geometries, minimizing a user-specified error function (mean absolute error, mean absolute percentage error, or mean squared error). This SLI can then be used to estimate distributions for other values which are only known at one of the geometry sets.

## Installation

You can install the development version of sliopt like so:

``` {r, eval=FALSE}
devtools::install_github("https://github.com/chris31415926535/sliopt")
```

## Overview of algorithm

The algorithm proceeds as follows:

* For each destination region $Bm$: 
  * For each origin region $Ai$ that intersects $Bm$:
    * Make note of the current SLI and the current overall SLI error value. 
    * Find each destination region $Bj$ such that $Aj$ and $Bj$ intersect, and the area of this intersection is greater than some minimal threshold proportion of $Ai$'s total area.
       * This is to exclude small overlaps that are due to data errors, for example if two regions are intended to share a border along a road but their shapefile representations overlap slightly along the border.
    * In turn, assign $Ai$ to each $Bj$ and calculate the resulting overall SLI error value If it is an improvement over the current SLI error value, update the SLI to assign $Ai$ to $Bj$ and make note of the improved error value.
  
By default the entire algorithm runs three times, with the destination regions $Bm$ shuffled to introduce some randomness.



## Example

This is a basic example using synthetic data.

First, we create two sets of geometries. The first set consists of two squares next to each other. The second set consists of rectangles that perfectly fill the set of squares. Note that the shapes are non-overlappingwithin each geometry set.

Next we create two synthetic population distributions. Using this point data, we can calculate "true" population counts over each geometry set.

Then, we can use those true values to create an optimized SLI.

The final image shows the smaller geometries shaded based on which of the larger two-square regions each is assigned to.

```{r example}
library(dplyr)
library(ggplot2)
library(sf)
library(sliopt)


layer1b <- list(list(matrix(c(0,0, 0,1, 1,1, 1,0, 0,0 ), byrow = TRUE, ncol = 2)),
                list(matrix(c(0,0, 0,1, -1,1, -1,0, 0,0 ), byrow = TRUE, ncol = 2))
)


layer2b <- list(list( matrix(c( -1,1, -1,.4, -.6,.4, -.6, 1, -1,1 ), byrow = TRUE, ncol = 2)),
                list( matrix(c( -1,0, .4, 0,  .4,.4,  -1,.4, -1,0 ), byrow = TRUE, ncol = 2)),
                list( matrix(c( .4,0,  1,0,   1, 1,  .4, 1, .4, 0), byrow = TRUE, ncol = 2)),
                list( matrix(c(-.6,1, -.6,.4 ,  .4, .4, .4, 1, -.6, 1), byrow = TRUE, ncol = 2))
)



layer1_sf <- lapply(layer1b, sf::st_polygon) %>%
  sf::st_sfc(crs="WGS84") %>%
  sf::st_sf() %>%
  dplyr::mutate(layer1_id = paste0("A", 1:nrow(.)))

layer2_sf <- lapply(layer2b, sf::st_polygon) %>%
  sf::st_sfc(crs="WGS84") %>%
  sf::st_sf() %>%
  dplyr::mutate(layer2_id = paste0("B", 1:nrow(.)))



# create two population clusters
pop_clusters <- create_pop_cluster(x=0.25, y=0.25, dx = 0.15, dy = 0.15, n=100, crs = "WGS84") %>%
  dplyr::bind_rows(create_pop_cluster(x=-0.75, y=0.75, dx = 0.25, dy = 0.25, n=100, crs = "WGS84"))

# create a plot
ggplot() +
  geom_sf(data = layer1_sf, colour = "red", size = 2) +
  geom_sf(data = layer2_sf, fill="lightblue", alpha = 0.6) +
  geom_sf(data = pop_clusters) +
  labs(title = "Point values over two overlapping sets of geometries")


layer1_count  <- sf::st_join(pop_clusters, layer1_sf) %>%
  dplyr::group_by(layer1_id) %>%
  dplyr::count(name = "layer1_n") %>%
  dplyr::filter(!is.na(layer1_id)) %>%
  sf::st_set_geometry(NULL) %>%
  dplyr::left_join(layer1_sf, ., by = "layer1_id")


layer2_count  <- sf::st_join(pop_clusters, layer2_sf) %>% 
  dplyr::group_by(layer2_id) %>%
  dplyr::count(name = "layer2_n") %>%
  dplyr::filter(!is.na(layer2_id)) %>%
  sf::st_set_geometry(NULL) %>%
  dplyr::left_join(layer2_sf, ., by = "layer2_id")

sli <- greedy_sli_search(to_shp = layer1_count, to_idcol = "layer1_id", to_valuecol = "layer1_n",
                  from_shp = layer2_count, from_idcol = "layer2_id", from_valuecol = "layer2_n", verbose = FALSE )

layer2_sf %>%
  left_join(sli, by = "layer2_id") %>%
  ggplot() +
  geom_sf(aes(fill = layer1_id)) +
  labs(title="Geometries shaded by SLI values")

```
